---
title: "AFD checklist"
author: "Payal Bal, Fonti Kar"
date: "2022-08-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r }
#install.packages("pacman")
pacman::p_load(dplyr, janitor, galah, stringr, tidyverse, arrow, stringr, data.table, CoordinateCleaner)

## Functions 
source("scripts/remove_improper_names_v2.R") # Version Fonti created
```

### Australian Faunal Directory taxonomy

```{r}
afd_species <- read_csv("output/afd_splist_full.csv")
all_species <- afd_species$VALID_NAME 
```

### Remove improper names 

```{r}
species_record <- remove_improper_names_v2(all_species,
                                           allow.higher.taxa = FALSE,
                                           allow.subspecies = TRUE)

# Double checking if there are any NAs in updated list
is.na(species_record$updated_list) |> table() 

# Filter records based on updated list
afd_species |> 
  dplyr::filter(VALID_NAME %in% species_record$updated_list) -> afd_species
```

### Remove invasive species 

The original workflow uses the GRIIS list downloaded from the ALA website. The code below uses a more up-to-date list from the ALA which is not currently on the website.

```{r}
# Read in the various txt files that make up the GRIIS list
distribution <- fread("data/dwca-griis-australia-v1.6/distribution.txt")
species <- fread("data/dwca-griis-australia-v1.6/speciesprofile.txt")
taxa <- fread("data/dwca-griis-australia-v1.6/taxon-edited.txt", fill = TRUE)

# Create the GRIIS list
griis_list <- taxa |>  
  full_join(distribution, by = "id") |>  
  full_join(species, by = "id")

# Use galah::search_taxa() to get tidy names and filter by noIssues
griis_ala_raw <- data.frame(search_taxa(griis_list$scientificName))

# Pre-process the GRIIS list and filter by noIssues and no missing values in species
griis_ala_tidy <- griis_ala_raw |>  
  filter(issues == "noIssue",
         !is.na(species)) |>  
  dplyr::select(search_term, taxon_concept_id, scientific_name, scientific_name_authorship)

# Cross matching AFD to updated GRIIS list and reporting
message(cat("number of AFD species listed in GRIIS: "),
        length(which(afd_species$VALID_NAME %in% griis_ala_tidy$scientific_name)))

message("AFD species in GRIIS - Global Register of Introduced and Invasive Species - Australia: ")

# Excluding from AFD list
message("Removing the following AFD species with GRIIS ...")
updated_griis_afd <- afd_species$VALID_NAME[which(afd_species$VALID_NAME %in% griis_ala_tidy$scientific_name)]

message("Updated AFD with species excluded ...")
afd_species <- afd_species[! afd_species$VALID_NAME %in% griis_ala_tidy$scientific_name, ]
```

#### Old workflow for removing invasive species

```{r ##}
griis_species <- read.csv("data/GRIIS_Global_Register_of_Introduced_and_Invasive_Species_Australia.csv")

message(cat("number of AFD species listed in GRIIS: "),
        length(which(afd_species$VALID_NAME %in% griis_species$Supplied.Name)))

message("AFD species in GRIIS - Global Register of Introduced and Invasive Species - Australia: ")
old_griis_afd <- afd_species$VALID_NAME[which(afd_species$VALID_NAME %in% griis_species$Supplied.Name)]

message("Removing the following AFD species with GRIIS ...")
afd_species <- afd_species[! afd_species$VALID_NAME %in% griis_species$Supplied.Name,]
```

#### Difference between found matches with updated list and old list

My recommendation is to use both lists for now and we can let Donald know that there. Under development here to explore differences between lists? 

```{r}
setdiff(updated_griis_afd, old_griis_afd) # Taxa captured in new list

setdiff(old_griis_afd, updated_griis_afd) # Taxa captured by old list but now in updated on 

data.frame(setdiff(old_griis_afd, updated_griis_afd)) %>% search_taxa() # Taxa in old list that are not in ALA

# Are any of species from old_griis_afd that is not in updated_griis_afd (i.e. not flagged in ALA search_taxa) found in the updated griis_list
# Get the genus
species <- paste(stringr::word(setdiff(old_griis_afd, updated_griis_afd), 2), collapse = "|")
genus <- paste(stringr::word(setdiff(old_griis_afd, updated_griis_afd), 1), collapse = "|")
wholenm <- paste(setdiff(old_griis_afd, updated_griis_afd), collapse = "|")

# Regex the genus to scientificName
str_subset(griis_list$scientificName, pattern = species) # Would these be synonms I wonder
str_subset(griis_list$scientificName, pattern = genus) # Would these be synonms I wonder
str_subset(griis_list$scientificName, pattern = wholenm) # Yes 2 full matches
```

## Remove Marine species

### Using World Register of Marine species

Passing taxonomic names to: `worrms::wm_records_names()` will return a dataframe of marine species! There is a limit of about 100 species in each API query so it can take a while! I have done this already and saved the output as a .csv. 

For more info: https://docs.ropensci.org/worrms/

You can request a download of WoRMS here: https://www.marinespecies.org/usersrequest.php. which they update once a month
Issue is there is n valid name just scientific name
We would need to make decisions about taxonomic status and do a pre-process ourselves. A very general clean gives me about  6,108 species - see workflow/worrms.rmd

- PB to export species list of semi-marine, marine/brackish to experts to cross verify this list

```{r}
worms <- read_csv("data/worms/worrms_marine_splist.csv")
names(worms)

## Any missing values in scientificname, valid_name?
is.na(worms$scientificname) |> table()
is.na(worms$valid_name) |> table()

## Do valid name match scientific name?
(worms$valid_name == worms$scientificname) |> janitor::tabyl()

## WoRMS has these really helpful identifiers of habitat occupancy of each taxa
## IA wants to exclude marine species exclusively so TRUE for isMarine and NA/FALSE for the others
worms %>% select(starts_with("is")) %>% colSums(., na.rm = TRUE)

## Number of strictly marine species
worms %>%
  filter(isMarine == 1,
         isBrackish == 0, isTerrestrial == 0, isFreshwater == 0, isExtinct == 0) -> marine_worms # 6768 species

## Which names don't match? I suggest cross-checking with AFD using both scientificname and valid_name for these marine species
marine_worms |> 
  filter(! valid_name == scientificname) |> 
  select(valid_name, scientificname) 

## First which species are excluded because they are marine 
afd_species |> 
  filter(VALID_NAME %in% marine_worms$valid_name) |>
  select(VALID_NAME) ->  excluded_mar_species
  
nrow(excluded_mar_species) # 5,850

## Exclude these species 
afd_species |> 
  filter(! VALID_NAME %in% marine_worms$valid_name) -> nomarine_nogriis_afd_species

# Checking
nrow(nomarine_nogriis_afd_species) + nrow(excluded_mar_species) == nrow(afd_species)

# Further exclusion using scientificname
## Which species are excluded by scientificname
nomarine_nogriis_afd_species |> 
  filter(VALID_NAME %in% marine_worms$scientificname) |> 
  select(VALID_NAME) -> excluded_mar_species_scientificname

nrow(excluded_mar_species_scientificname) # 575

## Further exclude excluded_mar_species_scientificname
nomarine_nogriis_afd_species |> 
  filter(! VALID_NAME %in% marine_worms$scientificname) -> nomarine_nogriis_afd_species_snm

nrow(nomarine_nogriis_afd_species_snm) + nrow(excluded_mar_species) + nrow(excluded_mar_species_scientificname) == nrow(afd_species)

### Reporting 
message(paste("Found ", nrow(excluded_mar_species), " marine species by validname in WoRMS"))
message(paste("After exclusion by validname, a further", nrow(excluded_mar_species_scientificname), "marine species were found by scientificname in WoRMS"))
message(paste("The total number of AFD species remaining following complete exclusion is ", nrow(nomarine_nogriis_afd_species_snm)))
```

## Identifying duplicates 

This section identify duplicates at the level of: 
- row
- `VALID_NAME`
- `COMPLETE_NAME`

The number of species > the number of _unique_ species implies that there are duplicates.
JM: use `COMPLETE_NAME` for finding duplicates


```{r}
## List duplicates comparing all columns
afd_species[duplicated(afd_species),] 

## Look for duplicates in specific columns
sum(is.na(afd_species$VALID_NAME)) # Count number of NA in VALID_NAME
length(afd_species$VALID_NAME) # Total species
length(unique(afd_species$VALID_NAME)) # Total unique VALID_NAME
length(unique(afd_species$COMPLETE_NAME)) # Total unique COMPLETE_NAME
```

## Duplicates in COMPLETE_NAME (excluding first appearance)

```{r}
message(cat("Number of duplicated COMPLETE_NAME (excluding first appearance): "), 
        length(afd_species$COMPLETE_NAME[duplicated(afd_species$COMPLETE_NAME)]))
message("duplicated COMPLETE_NAME: ")
afd_species$COMPLETE_NAME[duplicated(afd_species$COMPLETE_NAME)]
```

## Duplicates in COMPLETE_NAME (*including* first appearance)

```{r}

afd_species$COMPLETE_NAME[duplicated(afd_species$COMPLETE_NAME) | duplicated(afd_species$COMPLETE_NAME, fromLast=TRUE)] -> dup_temp

message(cat("#duplicates in COMPLETE_NAME (including first appearance) : ", length(dup_temp)))
message("duplicated COMPLETE_NAME: ")
dup_temp

# readr::write_csv(temp, "output/afd_completename_repeats.csv")

```

## Resolve duplicates from ALA list 

Removal or flagging of duplicates by `TAXON_GUID` 

ALA hasn't contributed here, these were completed *in consultation with JM*

```{r}
## using TAXON_GUID from afd_completename_repeats_JRM.csv
# afd_species <- unique(afd_species$COMPLETE_NAME)
removed_dups <- c("b05771ae-bda7-497a-87c4-b55a0ebc4ca1",
                  "03acc9d4-a209-4bf0-9972-bc7d35d56aea",
                  "83d18631-e160-42ad-8332-89e4b8ba82b6",
                  "c05506f8-0188-4850-8136-7b45ea35638e",
                  "0bb19498-874f-4c6c-a637-124ec9878130")

afd_species <- afd_species[which(afd_species$TAXON_GUID %!in% removed_dups),]
#readr::write_csv(afd_species, "./outputs/afd_species_clean.csv")
```

Looks like some duplicates but not all of them are removed 5/17

These are the ones excluded:

```{r}
afd_species |> filter(TAXON_GUID %in% removed_dups) |> pull(COMPLETE_NAME)

temp$COMPLETE_NAME

# Which of the 12 taxa are kept? 
dup_flagged <- afd_species[duplicated(afd_species$COMPLETE_NAME) | duplicated(afd_species$COMPLETE_NAME, fromLast=TRUE),]
dup_flagged %>% filter(! TAXON_GUID %in% removed_dups) %>% pull(COMPLETE_NAME)
```

## Various checks and tallies

Checks or tallies of species according to the number of words and particular formatting patterns in `VALID_NAME`

Note: '\' is an escape operator for various special characters

```{r}
## Counts for number of words
str_count(afd_species$VALID_NAME, pattern = "\\S+") |> janitor::tabyl()

## Counts for special characters in species names
stringr::str_count(afd_species$VALID_NAME, pattern = regex("\\\"")) |> sum() # Counting occurrences for \"
stringr::str_count(afd_species$VALID_NAME, pattern = regex("'")) |> sum() # Counting occurrences for '
stringr::str_count(afd_species$VALID_NAME, pattern = regex("\\(")) |> sum() 
stringr::str_count(afd_species$VALID_NAME, pattern = regex("\\[")) |> sum() 
```




