---
title: "MolluscaBase"
author: "Fonti Kar"
date: "2023-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, arrow, janitor, worrms, job, writexl, skimr)
```

[MolluscaBase](https://www.molluscabase.org/index.php) is a branch of the [World Register of Marine Species](https://www.marinespecies.org/) (WoRMs), this means that we can use WoRMS tools to get the data we need - score!


```{r}
# Read in AFD
afd_may <- read_csv_arrow("data/afd_May2023_clean.csv")
```

## MatchTool

MatchTool is available on [MolluscaBase](https://www.molluscabase.org/aphia.php?p=match) and [WoRMS](https://www.marinespecies.org/aphia.php?p=match), to match a user's taxonomic list to their database to verify any changes taxonomic updates/typos. 

This is a great method to shows what's in AFD and not in MolluscaBase/WoRMs, but first we need to prepare our data for these tools. 

### Filter out Molluscs from AFD

Here we are filtering out data for Molluscs only.

```{r}
# Filter to Molluscs and select the column we need
molluscs_afd <- afd_may |> filter(
  PHYLUM == "MOLLUSCA"
) |> 
  select(COMPLETE_NAME)

nrow(molluscs_afd)
print(molluscs_afd, n = 6)

# Save Mollusca
# write_csv(matchtool, "outputs/AFD_May_Molluscs.csv") 
```

Here we are creating our own scientific name and authority variable as this was not included as a separate column in the AFD dataset.

```{r}
# Split COMPLETE_NAME to create authority and scientific name
molluscs_afd_authority <- molluscs_afd |> 
  mutate(split = map(.x = COMPLETE_NAME,
                         ~str_split(.x, pattern = ", ")),
         scientific_name = map(split, 
                         ~pluck(.x, 1, 1)),
         authority = map(split, 
                         ~pluck(.x, 1, 2))) |> 
  select(-c(split, COMPLETE_NAME)) |> 
  unnest(cols = c("scientific_name","authority")) 

molluscs_afd_authority |> print(n = 6)
```

One limitation to the Matchtool on the website is that it can only accept files of 1499 rows (1500 technically but column headers is count as one!). So here we are splitting the dataset into chunks of 1499 rows and export as an Excel file (a hack from Frank Koehler so the commas in the authority doesn't affect the matches). 

```{r}
# Split so it has 1500 rows in total 
chunk <- 1499
n <- nrow(molluscs_afd_authority)
r  <- rep(1:ceiling(n/chunk),each=chunk)[1:n]
d <- split(molluscs_afd_authority,r)

d

# Assign the list with useful names
names(d) <- paste0("May_AFD_Mollusca_matchtool_", names(d))

names(d)

# Save each chunk as .csv
walk2(.x = d, .y = names(d),
     ~write_xlsx(.x, path = paste0("outputs/", .y, ".xlsx")))
```

Once I exported these files, we have to go to the [Matchtool on MB website]([MolluscaBase](https://www.molluscabase.org/aphia.php?p=match) ). 

I uploaded these and selected the following options: 

- Check Match authority
- Match up to ScientificName
- Check Match against MolluscaBase
- Selected authority in addition to defaults in Outputs

In the preview mode, the app shows you what variable your column headers are likely to be:

scientific_name ~ ScientificName
authority ~ ScientificNameAuthor

Matching takes a few moments once complete, you can save the output as.

If there are an ambiguous matches, the tool allows you to try clarify them by providing species name + authority to select from the drop down menu.

### Read in output from matchtool

```{r}
matched_output_1 <- read_xlsx("data/may_afd_mollusca_matchtool_1_matched.xlsx") |> 
  select(-3) |> 
  rename(afd_scientific_name = scientific_name)

matched_output_2 <- read_xlsx("data/may_afd_mollusca_matchtool_2_matched.xlsx") |> 
  select(-3) |> 
  rename(afd_scientific_name = scientific_name)

# Bind these together
mb_output_matchtools <- bind_rows(matched_output_1, matched_output_2) |> 
  clean_names()
```

### Investigate output from MatchTools

Match type categories: 

- `exact` all characters match exactly
- `exact_subgenus` an exact match, but including the subgenus
- `phonetic` sounds similar as, despite minor differences in spelling (soundex algorithm)
- `near_1` perfect match, except for one character. This is a quite reliable match
- `near_2` good match, except for two characters. This needs an extra check
- `near_3` good match, except for three characters. This definitely needs an extra check
- `match_quarantine` match with a name that is currently in quarantine. Any name that has been used in the - literature should in principle not be quarantined. So best to contact the WoRMS DMT about this
- `match_deleted` this is a match with a name that has been deleted and no alternative is available. Please contact the WoRMS DMT when you come across this.


```{r}
mb_output_matchtools |> 
  pull(match_type) |> 
  tabyl()
```

#### Taxa MolluscaBase can add

Send this to Frank Koehler

```{r}
mb_output_matchtools |> 
  filter(is.na(match_type)) 

mb_output_matchtools |> 
  filter(is.na(match_type)) |> 
  nrow()
```

#### Taxa AFD can correct

Send these to Matt and Hayley at AFD to correct

```{r}
mb_output_matchtools |> 
  filter(str_detect(match_type, "near")) 

mb_output_matchtools |> 
  filter(str_detect(match_type, "near"))  |> 
  nrow()
```

#### Phonetic taxa

Visually inspect these to see how close they are.
These look like typos to me by the AFD, needs to be checked by JM or JK
Once confirmed, send these to Matt and Hayley at AFD to correct

```{r}
mb_output_matchtools |> 
  filter(match_type == "phonetic")  |> 
  select(ends_with("name"))
```

#### Ambiguous taxa

These need to be followed up with JM/JK or Frank. 
Exporting these and putting into MatchTool again shows there are 2 possible `near_3` matches

```{r}
mb_output_matchtools |> 
  filter(match_type == "ambiguous") 

mb_output_matchtools |> 
  filter(match_type == "ambiguous") |> nrow()

mb_output_matchtools |> 
  filter(match_type == "ambiguous") |> 
  select(afd_scientific_name, authority) |> 
  write_xlsx("outputs/May_AFD_ambiguous_molluscs_matchtool.xlsx")
```

#### Try use WoRMs API for MatchTool

Fuzzy match = TRUE

```{r}
species_chunks <- split(match_clean$scientific_name, ceiling(seq_along(match_clean$scientific_name)/120))


# Testing entire species list
job({
output <- map(match_clean$scientific_name,
    possibly(~wm_records_name(.x) |> 
               mutate(search_term = .x))) |> 
  discard(.p = ~is.null(.x)) 
})


output_df <- output |> list_rbind()

# write_csv(output_df, "outputs/API_WoRMS_matched.csv")
matchtool_fuzzy <- read_csv("outputs/API_WoRMS_matched.csv")
```

```{r}
matchtool_fuzzy |> 
  pull(match_type) |> 
  tabyl()

output_df |> 
  mutate(match = search_term == scientificname) |> 
  select(AphiaID:url,match, match_type, search_term, scientificname:modified) |>
  filter(match == FALSE)
```

Fuzzy match != TRUE

```{r}
# Testing entire species list
job({
output_nofuzz <- map(match_clean$scientific_name,
    possibly(~wm_records_name(.x, fuzzy = FALSE) |> 
               mutate(search_term = .x))) |> 
  discard(.p = ~is.null(.x)) 
})


output_df <- output |> list_rbind()

# write_csv(output_df, "outputs/API_WoRMS_matched.csv")
no_fuzzy_matchtool <- read_csv("outputs/API_WoRMS_matched.csv")
```

## Spotting typos in the AFD

Try `wm_records_taxamatch()`

There are several types of matches:

- `exact` all characters match exactly
- `exact_subgenus` an exact match, but including the subgenus
- `phonetic` sounds similar as, despite minor differences in spelling (soundex algorithm)
- `near_1` perfect match, except for one character. This is a quite reliable match
- `near_2` good match, except for two characters. This needs an extra check
- `near_3` good match, except for three characters. This definitely needs an extra check
- `match_quarantine` match with a name that is currently in quarantine. Any name that has been used in the - literature should in principle not be quarantined. So best to contact the WoRMS DMT about this
- `match_deleted` this is a match with a name that has been deleted and no alternative is available. Please contact the WoRMS DMT when you come across this.

```{r}
# Job takes 2hrs and 40 mins to run  

# job({
#   output_df <- map(match_clean$scientific_name,
#                    possibly(~wm_records_taxamatch(.x, marine = FALSE)  |>
#                               pluck(1)  |>
#                               mutate(search_term = .x) |>
#                               discard(.p = ~is.null(.x))
#                    )
#   ) |> 
#     list_rbind()
#   
#   saveRDS(output_df, "outputs/worrms_taxamatch")
# })

worrms_taxamatch <- readRDS("outputs/worrms_taxamatch")

# Types of fuzzy match
worrms_taxamatch |> 
  pull(match_type) |> 
  tabyl()

# Which are the taxa that were not matched in WoRMS (i.e AFD has a name that WoRMS does not)
setdiff(match_clean$scientific_name, worrms_taxamatch$search_term) |> length()

# Which ones that are not exact matches - this maybe ones that we need to update OR we have a newer name or completely unrelated species
worrms_taxamatch |> 
  filter(! match_type == "exact") |> 
  select(search_term, ends_with("name"),  match_type, valid_authority) |> 
  arrange(match_type) |> 
  print(n = 100) |> 
  write_csv(paste0("outputs/AFD_WoRMS_taxamatch_notexactmatchtype", Sys.Date(), "_.csv"))
```

## Read in data sent by Data Management team for all of Australia

[Distribution tool](https://www.marinespecies.org/aphia.php?p=checklist). 

Select Geounit: "Australia (Nation)"
Type: specimen
Rank: lower or equal to Mollusca
Extant only: TRUE
Marine: FALSE
Distribution status: Valid
Synonyms: Only accepted names

Note: Manual download is capped at 1000 taxa so we have to email the [WoRMS Data Management Team (DMT](mailto::info@marinespecies.org) to request for a Excel Spreadsheet download. 
There is currently no programmatic way to achieve this unfortunately. DMT sends a version where: 
Select Geounit: "Australia (Nation)", Type: specimen, Rank: lower or equal to Mollusca, Synonyms, Sort by synonyms

The DMT pointed to this [Shiny app](http://rshiny.vsc.vliz.be:52038/standardized_distributions_test/) to get distribution related data, but unfortunately only works for marine 'zones'. 


```{r}
aus_molluscs <- readxl::read_xlsx("data/WoRMS_Mollusca_australia_20230621.xlsx")

skim(aus_molluscs)
```

### Obtainx synonyms WoRMs synonyms API

API returns synoynms by AphiaID, valid_AphiaID is the currently accepted one
s
```{r}
# job::job({aus_moll_syn_check <- map(test,
#                                     possibly(~wm_synonyms(.x) |>
#                                                mutate(search_AphiaID = .x))
#                                     )
# 
# aus_moll_syn_check_df <- aus_moll_syn_check |> 
# discard(.p = ~is.null(.x)) |> 
#   bind_rows()
# 
# saveRDS(aus_moll_syn_check_df, "outputs/Worms_Aus_synonyms")
# })

aus_moll_syn_check <- readRDS("outputs/Worms_Aus_synonyms")

# Select synonym ids
syns <- aus_moll_syn_check |> 
  pull(valid_AphiaID) |> 
  unique()

# Display synonyms
aus_molluscs |> 
  filter(AphiaID %in% syns) 

# Exclude synonyms
aus_molluscs_nosyns <- aus_molluscs |> 
  filter(! AphiaID %in% syns)
```

### Obtain more detailed records for remaining taxa

Using the `wm_records_taxamatch` API. We can retrieve for indepth variables (habitat, taxonomy status) for further exclusions. 

```{r}
# job({
#   output_df <- map(unique(aus_molluscs_valid$ScientificName),
#                    possibly(~wm_records_taxamatch(.x, marine = FALSE)  |>
#                               pluck(1)  |>
#                               mutate(search_term = .x) |>
#                               discard(.p = ~is.null(.x))
#                    )
#   ) |>
#     list_rbind()
# 
#   saveRDS(output_df, "outputs/worrms_Australia_taxamatch_remove_marine")
# })

output_df <- readRDS("outputs/worrms_Australia_taxamatch_remove_marine")

# Checking number of rows
length(aus_molluscs_valid$ScientificName) # ] 5686
length(unique(aus_molluscs_valid$ScientificName)) # 5349
nrow(output_df) # 5866
```

## Duplicated records

```{r}
# What is duplicated
output_df |> 
  filter(duplicated(search_term)) |> 
  nrow() 
```

## Unaccepted names

```{r}
# Exclude unaccepted names
output_df |> 
  pull(status) |> 
  tabyl()

accepted_molluscs <- output_df |> 
  filter(status == "accepted")
```

## Extinct taxa

```{r}
# Exclude extinct taxa
accepted_noextinct_molluscs <- accepted_molluscs |> 
  filter(!isExtinct == TRUE) 
```

## Higher taxonomy

Taxon that cannot be identified down to species is not useful for our backbone

```{r}
# Exclude higher taxon rank
accepted_noextinct_molluscs |> 
  pull(rank) |> 
  tabyl()

# Fix NA in rank
accepted_noextinct_molluscs |> 
  mutate(rank2 = case_when(is.na(rank) ~ "Species",
                          TRUE ~ rank))  |> 
  pull(rank) |> 
  tabyl()

accepted_noextinct_molluscs |> 
  mutate(rank2 = case_when(is.na(rank) ~ "Species",
                          TRUE ~ rank))  |> 
  pull(rank2) |>  # Can see no NA
  tabyl()

accepted_noextinct_molluscs_rankfixed <- accepted_noextinct_molluscs |> 
  mutate(rank_fixed = case_when(is.na(rank) ~ "Species",
                          TRUE ~ rank))

# Exclude higher taxon 
accepted_noextinct_molluscs_rankfixed |> 
  pull(rank_fixed) |> 
  tabyl()

accepted_noextinct_molluscs_sp <- accepted_noextinct_molluscs_rankfixed |> 
  filter(!rank_fixed == "Genus")
```

### Excluding Marine taxa

There are habitat flags in WoRMS that can tell us whether a species is marine/brackish/terrestrial/freshwater or a combination of these. 

NA in habitat flags represents missing information and should not be treated as a '0'

Criterion on what data we want to keep:

- Terrestrial taxa (`isTerrestrial` == 1)
- Freshwater taxa (`isFreshwater` == 1)
- Brackish taxa (`isBrackish` == 1)
- Any combination of the above (T/F/W) that are also found in marine environments (e.g. `isMarine` == 1 & `isBrackish` == 1)
- Any combination of the above (T/F/W) and their status in marine environment is unknown (e.g. `isMarine` = NA & `isFreshwater` == 1)

```{r}
nrow(accepted_noextinct_molluscs_sp)

# Not marine or missing info in Marine
accepted_noextinct_molluscs_sp |> 
  filter(isMarine == 0 | is.na(isMarine)) |> 
  distinct() |> 
  nrow() # 118

# Taxon found in at least one habitat that we want to keep
# This is what we want
accepted_noextinct_molluscs_sp |> 
  filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
  distinct() |> 
  nrow() #119

accepted_noextinct_molluscs_sp |> 
  filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
  distinct() |> 
  View()

# Difference
setdiff(accepted_noextinct_molluscs_sp |> 
          filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
          distinct() |> 
          pull(scientificname),
        accepted_noextinct_molluscs_sp |> 
          filter(isMarine == 0 | is.na(isMarine)) |> 
          distinct() |> 
          pull(scientificname))

# Found in three habitats
accepted_noextinct_molluscs_sp |> 
  filter(scientificname == "Paromoionchis tumidus") |> 
  select(scientificname, search_term, starts_with("is"))  

# Check: Same as above
# Covered by Taxon found in at least one habitat that we want to keep
accepted_noextinct_molluscs_sp |> 
  filter(isMarine == TRUE & isTerrestrial == TRUE | isMarine == TRUE & isFreshwater == TRUE | isMarine == TRUE & isBrackish == TRUE) |> 
  select(scientificname, search_term, starts_with("is"))  

# Strictly marines - to exclude
accepted_noextinct_molluscs_sp |> 
  filter(isMarine == TRUE & isTerrestrial == FALSE & isFreshwater == FALSE & isBrackish == FALSE) |> 
  distinct() |> 
    nrow()
  
accepted_noextinct_molluscs_sp |> 
  filter(isMarine == TRUE & isTerrestrial == FALSE & isFreshwater == FALSE & isBrackish == FALSE) |> 
  distinct() |> 
  View()

strictly_marines <- accepted_noextinct_molluscs_sp |> 
  filter(isMarine == TRUE & isTerrestrial == FALSE & isFreshwater == FALSE & isBrackish == FALSE) |> 
  distinct() |> 
  pull(scientificname)

# Excluding strictly marines should equal to Taxon found in at least one habitat that we want to keep
accepted_noextinct_molluscs_sp |> 
  filter(! scientificname %in% strictly_marines) |> 
  distinct() |> 
  nrow() #130

# Difference
diff_taxa <- setdiff(accepted_noextinct_molluscs_sp |> 
          filter(! scientificname %in% strictly_marines) |> 
          distinct() |> 
          pull(scientificname),
        accepted_noextinct_molluscs_sp |> 
          filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
          distinct() |> 
          pull(scientificname))

# These are taxa where they are marine but unknown for other fields
accepted_noextinct_molluscs_sp |> 
  filter(scientificname %in% diff_taxa) |> 
  select(scientificname, search_term, starts_with("is"))  

accepted_noextinct_molluscs_sp |> 
  filter(!scientificname %in% strictly_marines) |> 
  filter(!is.na(isBrackish))  |> 
  select(scientificname, search_term, starts_with("is")) |> 
  distinct() # BAM this matches with taxon above found in at least one habitat that we want to keep

# BAM this matches with taxon above found in at least one habitat that we want to
setdiff(accepted_noextinct_molluscs_sp |> 
          filter(!scientificname %in% strictly_marines) |> 
          filter(!is.na(isBrackish))  |> 
          select(scientificname, search_term, starts_with("is")) |> 
          distinct() |> 
          pull(scientificname),
        accepted_noextinct_molluscs_sp |> 
          filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
          distinct() |> 
          pull(scientificname)
)

setdiff(
        accepted_noextinct_molluscs_sp |> 
          filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
          distinct() |> 
          pull(scientificname),
        accepted_noextinct_molluscs_sp |> 
          filter(!scientificname %in% strictly_marines) |> 
          filter(!is.na(isBrackish))  |> 
          select(scientificname, search_term, starts_with("is")) |> 
          distinct() |> 
          pull(scientificname)
)


accepted_keep_taxa <- accepted_noextinct_molluscs_sp |> 
  filter(isTerrestrial == TRUE | isFreshwater == TRUE | isBrackish == TRUE) |> 
  distinct() 

# Cool code
# accepted_noextinct_molluscs_sp |> 
#   filter(isMarine == FALSE | if_any(isBrackish:isFreshwater, ~ . == 1)) |> 
#   select(scientificname, search_term, starts_with("is"))  |> 
#   distinct()
```

## Matches to AFD
```{r}
matches <- intersect(accepted_keep_taxa
 |> pull(scientificname), afd_may |> pull(FULL_NAME))

matches
```

## New taxon additions

```{r}
new <- setdiff(accepted_keep_taxa |> pull(scientificname), afd_may |> pull(FULL_NAME))

new
```

```{r}
# Checks
length(matches) + length(new)
```

```{r}
# Cross compare with Taxamatch data
setdiff(new, unique(worrms_taxamatch$scientificname)) |> length()
setdiff(new, unique(match_clean$scientific_name)) |> length()

# Locality breakdown
aus_molluscs |> tabyl(Locality)
```

