---
title: "Cleaning the World Spider Catalog List"
author: "Fonti Kar"
date: "2023-05-08"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      include=FALSE)
```

# Cleaning the World Spider Catalog

Data can be accessed via the [WSC Data resources](https://wsc.nmbe.ch/dataresources) page. Bulk download is from their "Daily Species Exports" heading

## Load dependencies 

```{r, include=TRUE}
pacman::p_load(tidyverse, skimr, arrow, janitor, worrms, here, job)
```

## Read in data

```{r, include=TRUE}
wsc <- read_csv(here("data/WSC_taxonomic_checklist.csv"))
```

## Getting familiar with data

```{r, include=TRUE}
skim(wsc)
```

## What are the relevant levels of 'distribution' 

```{r}
wsc |> 
  pull(distribution) |> str_subset(pattern = "Australia") |> 
  unique()

# Pattern to ignore Introduced to Australia
wsc |> 
  pull(distribution) |> str_subset(pattern = "Introduced to Australia") 
```


## Precleaning 

No scientific_name found in dataset. Here, we created a `scientific_name` variable

```{r, include=TRUE}
wsc_scnm <- wsc |> 
  mutate(scientific_name = case_when(
    is.na(subspecies) ~ paste(genus, species),
    !is.na(subspecies) ~ paste(genus, species, subspecies)
  )
  ) |> 
  select(speciesId:species_lsid, scientific_name, family:distribution)

wsc_scnm
```

## Remove non-Australian species

"Introduced to Australia" described the introduction of a foreign species to Australia (exclude these)
"Country to Australia" e.g. Indonesia to Australia is describing the range of the taxa. 

```{r, include=TRUE}
# All records that are flagged Australia
oz_wsc <- wsc_scnm |> 
  filter(str_detect(distribution, "Australia"))

# Use regular expression to detect Introduced to Australia
# These are the ones that we will exclude
oz_wsc |> 
  pull(distribution) |> 
  str_subset("Introduced to.*Australia")  |> # .* matches instances where there are as many spaces/characters between "Introduced to" and "Australia. E.g. Introduced to St. Helena, Australia, New Zealand
  head()

# Number to be excluded
introduced_taxa <- oz_wsc |> 
  pull(distribution) |> 
  str_subset("Introduced to.*Australia") |> 
  length()

# Use regular expression to detect and exclude "Introduced to Australia"
oz_wsc_nointroduced <- oz_wsc |> 
  filter(! str_detect(distribution, "Introduced to.*Australia"))

nrow(oz_wsc) - introduced_taxa == nrow(oz_wsc_nointroduced)
```

Exclude `r introduced_taxa` taxa that were Introduced to Australia. `r nrow(oz_wsc_nointroduced)` remaining.

## Remove improper names

```{r, echo=TRUE}
source(here("scripts/remove_improper_names_v2.R"))

oz_wsc_clean_names <- oz_wsc_nointroduced |> 
  pull(scientific_name) |> 
  remove_improper_names_v2()

oz_wsc_clean_names$improper_species
```

There were `r length(oz_wsc_clean_names$improper_species)` improper species.

## Remove marine species

```{r}
# job({
# wm_oz_wsc <- map(oz_wsc_nointroduced |> pull(scientific_name),
#     possibly(~wm_records_name(.x, fuzzy = FALSE) |>
#                mutate(search_term = .x))) |>
#   discard(.p = ~is.null(.x)) |>
#   bind_rows()
# 
# 
# saveRDS(wm_oz_wsc, here("outputs/WSC_worms_marine_check"))
# })

wm_oz_wsc <- readRDS(here("outputs/WSC_worms_marine_check"))
```

`r nrow(wm_oz_wsc)` marine species found in WSC data.

```{r, include=TRUE}
wm_oz_wsc |> select(search_term, scientificname, valid_name, AphiaID, isMarine)

# Exclude these from WSC checklist
oz_wsc_nointroduced_nomarine <- oz_wsc_nointroduced |> 
  filter(! scientific_name %in% wm_oz_wsc$search_term)
```

## Duplicates

```{r, include=TRUE}
duplicated_taxa <- oz_wsc_nointroduced_nomarine |> 
  filter(duplicated(scientific_name)) |> 
  pull(scientific_name)
```

We found `r length(duplicated_taxa)` duplicated taxa. Author, Year and Distribution appears to differ between the two rows. **NO ACTION YET** - Keep both or which to keep?

```{r, include=TRUE}
oz_wsc_nointroduced_nomarine |> 
  filter(scientific_name == duplicated_taxa) 
```

## Matches to AFD

```{r}
afd_may <- read_csv(here("data/afd_May2023_clean.csv"))

matches <- intersect(oz_wsc_nointroduced_nomarine |> pull(scientific_name), afd_may |> pull(FULL_NAME))
```

There were `r length(matches)` matches found between Australian taxa in the WSC and the AFD. These include (6 examples show here):

```{r, include=TRUE}
matches |> 
  head()
```

## New taxon additions

```{r}
new <- setdiff(oz_wsc_nointroduced_nomarine |> pull(scientific_name), afd_may |> pull(FULL_NAME))
```

There were `r length(new)` *new taxa found* in the WSC that we can add to the AFD. These include: 

```{r, include=TRUE}
sort(new)
```

## Flag in AFD

```{r, include=TRUE}
# Select relevant columns and case column case to match AFD and add "source" variable to flag new taxa comes from WSC
new_taxa <- oz_wsc_nointroduced_nomarine |> 
  filter(scientific_name %in% new) |> 
  select(scientific_name, family, genus, species , author, year) |> 
  mutate(source = "WSC",
         year = as.character(year)) |> 
  clean_names(case = "all_caps") 

# Check
updated_afd_wsc <- bind_rows(afd_may |> mutate(SOURCE = "AFD"),
          new_taxa)  

updated_afd_wsc$SOURCE |> tabyl()
```

## Final checks

```{r, include=TRUE}
length(matches) + length(new) + length(duplicated_taxa) == nrow(oz_wsc_nointroduced_nomarine)  
```

